Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-08-22 Multiresolution_framework/ltx/response_to_reviews.tex

        \item Section 6 also borrows heavily from Dewar et al, e.g. cf equation (58) of this paper with equation (19) of Dewar et al: the state transition matrix is a perturbation of the corresponding matrix used in that earlier paper. The approach used in both papers seems to be the same with an RTS smoother used in the E-step (this was also used and explained in Freestone et al.) and an M-step optimizing a cost function, quadratic in the parameters, cf. equation (62) with equation (29) of Theorem 2 of Dewar et al. In general, the derivations of section 6 and appendix B follow closely or are very similar to those of Dewar et al.

		\emph{To clarify the advantage of the M-step over the implementation in Dewar et al. we have added a paragraph to Section 8.1 (now 8.2). Further comments for the reviewer are added below.}

\emph{The inverse framework in Dewar et al. (2009) is based on a general integro-differential equation model, which is not suitable when neural field models are considered. As a result, the necessary modifications have been made. }

\emph{This paper has a novel implementation of the M-step, which significantly simplifies the computational complexity of the EM-based algorithm proposed by Dewar et al. (2009). These modifications are significant and crucial for high-dimensional systems. More specifically, implementing the multi-resolution framework introduced in the paper would be impossible without this new contribution. The differences are highlighted in Table.~4 of the manuscript, namely from $O(n_x^4n_{\theta}^2)$ to $O(n_x^2n_{\theta}^2)$, see equation (64) of this paper and equation (31) of Dewar et al. (2009), where the static and dynamic components are separated by defining $\boldsymbol \Gamma_2$ and $\mathbf U$ (see equations (66) and (67)). A simplification is also performed to calculate $\boldsymbol\upsilon_0, \boldsymbol\upsilon_1$ (see equations (62) and (63) of this paper and equation (30) of Dewar et al. (2009)).}

@@

		\item Frankly, I would off-load all of the very difficult mathematics to the appendices and supplementary data, and focus a NeuroImage paper on the results of image analysis.

			\emph{We attempted to move the derivation from Eq. 22 to Eq. 33 to the appendix. This led to more complications, where it was very difficult to define the various components of the model and a notational nightmare. To guide the reader who may just be interested in the conceptual framework and results we have added ``The conceptual framework and the key results can be interpreted without following this derivation in detail.'' below Eq. 22.}

			\item I would also recommend archiving code that, for instance generates Figure 11, so that the readership has a chance to replicate these findings, and adapt this method to their own data.

 			\emph{We have uploaded the Python source codes for our numerical simulations as supplementary material.}

@@

			\item Between Eq. 19 and 20 change
			`write the neural field and connectivity kernel decomposition' To `write the connectivity kernel and neural field decomposition'
			To maintain order with Eqs. 20 and 21.

			\emph{We have changed the text to `the connectivity kernel and neural field decomposition' to maintain consistency with Eqs. 20 and 21.}

			 \end{enumerate}  
			\subsection{Minor Comments}
			\begin{enumerate}
			 \item 	Define `one-off' on pg. 16

 			\emph{We have clarified this point by changing the text to `All $n_x^2$ blocks of $\mathbf U$, $\boldsymbol\Gamma_1$ and $\boldsymbol\Gamma_2$ can be computed and stored prior to the commencement of the EM iterations' }

### 2012-08-22 Multiresolution_framework/ltx/manuscript.tex
comment for tight (frame) and redundancy

The state-transition equation can be simplified if the chosen family of scaling and wavelet functions are isotropic where
\begin{equation}
	\boldsymbol{\lambda}_{i}(r-r') = \boldsymbol{\lambda}(2\mathbf{c}_{i}+r'-r), 
\end{equation}
where $c_i$ is the centre of the $i^{th}$ basis function of the kernel decomposition. Following this, we can switch the order of operations with the inner product and convolution in Eq.~\eqref{eq:ApproxDiscreteTimeModel4}. With this switch the convolution is now between the scaling functions and wavelets for the field and the scaling functions and wavelets for the connectivity kernel. This convolution can be calculated analytically when using B-spline scaling and wavelet functions and can be predefined, since these are static functions. This has major computational advantages when solving the estimation problem. %\parham{Although this switch holds theoretically it should be noted that it will introduce numerical error, specially close to the edges, due to the finite extent of the field. } 
The predefined convolution can be written as
\begin{equation}\label{eq:Gammaij}
	\left[\Gamma(r')\right]_{:i} = T_s \mathbf{\Lambda}_{x}^{-1}\int_{\Omega} \boldsymbol\mu\left(r\right)\boldsymbol\lambda\left(2c_{i} + r'-r\right) \,\mathrm{d}r,
\end{equation} 
 where $\left[\Gamma(r')\right]_{:i}$ denotes the $i^{th}$ column of  $\left[\Gamma(r')\right]$ giving
\begin{equation}
	\mathbf{x}_{t+1} = 
	\xi \mathbf{x}_t + 
	\int_{\Omega} \Gamma\left(r'\right)f\left(\boldsymbol\mu^\top\left(r'\right) \mathbf{x}_t\right) 
	\, \mathrm{d}r' \boldsymbol\theta
	+ \mathbf w_t.
\end{equation}
The observation equation of the state-space model is found by substituting decomposition \eqref{eq:FieldFiniteExpansion}
 into Eq.~\eqref{eq:ObservationEquation} giving
\begin{equation}\label{eq:ReducedObservationEquation} 
	\mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\epsilon}_t,
\end{equation}
where each element of the observation matrix, $\mathbf{C}$, is given by
\begin{equation}\label{eq:Observationmatrix}
	\mathbf{C}_{ik} \triangleq \int_{\Omega}m(r_i -r')\boldsymbol{\mu}_k(r') \, \mathrm{d}r'.
\end{equation}
\section{B-Spline MRAIDE in State-Space }
\subsection{A Brief Introduction to B-Spline Wavelets}
This section is included to provide the interested reader with an overview of the theory used in constructing the wavelets that form the basis of the multi-resolution model. Although a deep understanding of the theory presented in this section is not required for an understanding of the key results of the paper, it serves as a reference for implementing the methods.

@@

All $n_x^2$ blocks of $\mathbf U$, $\boldsymbol\Gamma_1$ and $\boldsymbol\Gamma_2$  can be computed as a one-off before the commencement of the EM iterations (Step~1 in Table~\ref{alg:EMsteps} ), which increases the speed of the M-step significantly compared to the implementation of \citet{Dewar2009}. The complexity of each step of the algorithm is summarized in Table~\ref{table:MRA-ComputationalComplexity} where for ease of comparison the computational complexity of the method of \citet{Dewar2009} is also identified. It can be seen from the table that the state dimension is an important factor at each iteration of the estimation algorithm, emphasizing the advantage of the proposed algorithm compared to the method proposed in \citet{Dewar2009}. This also suggests the use of an efficient algorithm to choose the minimum number of basis functions (states of the system) while achieving the desired details of the underlying field.  

@@

To demonstrate the performance of the MRAIDE estimation framework, data was generated synthetically \parham{using Amari neural field model  (Eqs.~\eqref{eq:DiscreteTimeModel} and \eqref{eq:ObservationEquation})}, allowing a comparison between true and estimated parameters \footnote{\parham{The Python source code for our numerical simulations is available from the authors on request.}}. In doing so, the membrane time constant was set to $\tau = 10$~ms \citep{David2003}, and the sampling time was chosen ten times faster \citep{Stephan2008}, i.e. $T_s = 1$~ms, ensuring the discrepancy between the continuous neural field model and its discrete approximation is small. The connectivity kernel used for data generation composed of two B-spline scaling functions at $j=0$ and $j=1$ forming a compact support Mexican hat shape 

@@

% This is depicted in Fig.~\ref{fig:KernelEstimate}(A).

Two different experiments were designed to test the framework. In the first one, 200 realizations of 1 second of data were used, where the field disturbance, $e_t$, and the observation noise, $\boldsymbol\epsilon_t$, were regenerated each run. This way the net shape of the connectivity kernel could be formed by averaging over the parameters estimates from each realization. In the second experiment, a single data set was generated and the field estimations are compared for five state-space models, each accounting up to a certain level of approximation, i.e. $j=0$ to $j=4$. \parham{We show it is possible to determine the level of approximation to accurately estimate the underlying field prior to the EM-based algorithm.} All parameters for the model are given in Table~\ref{tab:Parameters}.

@@

In order to apply estimation algorithms to neural field models a finite-dimensional state-space model must be formed. The problem of representing a continuous field with a finite number of states is solved by the basis function decomposition. An alternative to using the B-spline wavelets are Gaussian basis functions, which were adopted by \cite{Freestone2011} to decompose the field and connectivity kernel. One clear advantage of B-spline scaling and wavelet functions is the MRA capability, allowing the simultaneous representation of the neural field at different spatial scales. B-spline scaling and wavelet functions are compactly supported, eliminating  numerical errors caused by the Gaussian basis function. The partition of unity (ability to represent a zero spatial frequency field) of these basis functions enable the reconstruction of the neural field when it saturates. This may be beneficial to model neurological disorders such as epilepsy. Attaining fine-scale approximation of the neural field using Gaussian basis functions results in an ill-posed problem as the number of states is increased significantly. However, this is not the case when using B-spline scaling and wavelet functions due to their semi-orthogonality property.\\ \citet{Freestone2011} showed experimentally that Gaussian functions are a suitable choice for modeling the observation kernel when intracranial measurements are considered. B-splines converge to Gaussian functions as the order of the B-spline, $m$, tends to infinity. The cubic B-spline is already a very close approximation to the Gaussian function. Therefore, these should also be suitable for modeling the observation kernel while having the advantage of being compactly supported.

@@

\dean{The basis structure (i.e., non-redundant versus a (tight) frame representation) is a requirement for this framework.}

>>>>

The state-transition equation can be simplified if the chosen family of scaling and wavelet functions are isotropic where
\begin{equation}
	\boldsymbol{\lambda}_{i}(r-r') = \boldsymbol{\lambda}(2\mathbf{c}_{i}+r'-r), 
\end{equation}
where $c_i$ is the centre of the $i^{th}$ basis function of the kernel decomposition. Following this, we can switch the order of operations with the inner product and convolution in Eq.~\eqref{eq:ApproxDiscreteTimeModel4}. With this switch the convolution is now between the scaling functions and wavelets for the field and the scaling functions and wavelets for the connectivity kernel. This convolution can be calculated analytically when using B-spline scaling and wavelet functions and can be predefined, since these are static functions. This has major computational advantages when solving the estimation problem.  
The predefined convolution can be written as
\begin{equation}\label{eq:Gammaij}
	\left[\Gamma(r')\right]_{:i} = T_s \mathbf{\Lambda}_{x}^{-1}\int_{\Omega} \boldsymbol\mu\left(r\right)\boldsymbol\lambda\left(2c_{i} + r'-r\right) \,\mathrm{d}r,
\end{equation} 
 where $\left[\Gamma(r')\right]_{:i}$ denotes the $i^{th}$ column of  $\left[\Gamma(r')\right]$ giving
\begin{equation}
	\mathbf{x}_{t+1} = 
	\xi \mathbf{x}_t + 
	\int_{\Omega} \Gamma\left(r'\right)f\left(\boldsymbol\mu^\top\left(r'\right) \mathbf{x}_t\right) 
	\, \mathrm{d}r' \boldsymbol\theta
	+ \mathbf w_t.
\end{equation}
The observation equation of the state-space model is found by substituting decomposition \eqref{eq:FieldFiniteExpansion}
 into Eq.~\eqref{eq:ObservationEquation} giving
\begin{equation}\label{eq:ReducedObservationEquation} 
	\mathbf{y}_t = \mathbf{C}\mathbf{x}_t + \boldsymbol{\epsilon}_t,
\end{equation}
where each element of the observation matrix, $\mathbf{C}$, is given by
\begin{equation}\label{eq:Observationmatrix}
	\mathbf{C}_{ik} \triangleq \int_{\Omega}m(r_i -r')\boldsymbol{\mu}_k(r') \, \mathrm{d}r'.
\end{equation}
\section{B-Spline MRAIDE in State-Space }
\subsection{A Brief Introduction to B-Spline Wavelets}
This section is included to provide the interested reader with an overview of the theory used in constructing the wavelets that form the basis of the multi-resolution model. Although a deep understanding of the theory presented in this section is not required for an understanding of the key results of the paper, it serves as a reference for implementing the methods.

@@

All $n_x^2$ blocks of $\mathbf U$, $\boldsymbol\Gamma_1$ and $\boldsymbol\Gamma_2$  can be computed and stored prior to the commencement of the EM iterations (Step~1 in Table~\ref{alg:EMsteps} ), which increases the speed of the M-step significantly compared to the implementation of \citet{Dewar2009}. The complexity of each step of the algorithm is summarized in Table~\ref{table:MRA-ComputationalComplexity} where for ease of comparison the computational complexity of the method of \citet{Dewar2009} is also identified. It can be seen from the table that the state dimension is an important factor at each iteration of the estimation algorithm, emphasizing the advantage of the proposed algorithm compared to the method proposed in \citet{Dewar2009}. This also suggests the use of an efficient algorithm to choose the minimum number of basis functions (states of the system) while achieving the desired details of the underlying field.  

@@

To demonstrate the performance of the MRAIDE estimation framework, data was generated synthetically using Amari neural field model  (Eqs.~\eqref{eq:DiscreteTimeModel} and \eqref{eq:ObservationEquation}), allowing a comparison between true and estimated parameters. In doing so, the membrane time constant was set to $\tau = 10$~ms \citep{David2003}, and the sampling time was chosen ten times faster \citep{Stephan2008}, i.e. $T_s = 1$~ms, ensuring the discrepancy between the continuous neural field model and its discrete approximation is small. The connectivity kernel used for data generation composed of two B-spline scaling functions at $j=0$ and $j=1$ forming a compact support Mexican hat shape 

@@

Two different experiments were designed to test the framework. In the first one, 200 realizations of 1 second of data were used, where the field disturbance, $e_t$, and the observation noise, $\boldsymbol\epsilon_t$, were regenerated each run. This way the net shape of the connectivity kernel could be formed by averaging over the parameters estimates from each realization. In the second experiment, a single data set was generated and the field estimations are compared for five state-space models, each accounting up to a certain level of approximation, i.e. $j=0$ to $j=4$. We show it is possible to determine the level of approximation to accurately estimate the underlying field prior to the EM-based algorithm. All parameters for the model are given in Table~\ref{tab:Parameters}.

@@

In order to apply estimation algorithms to neural field models a finite-dimensional state-space model must be formed. The problem of representing a continuous field with a finite number of states is solved by the basis function decomposition. The basis structure (i.e., non-redundant versus a (tight) frame representation) is a requirement for the multi-resolution framework. An alternative to using the B-spline wavelets are Gaussian basis functions, which were adopted by \cite{Freestone2011} to decompose the field and connectivity kernel. One clear advantage of B-spline scaling and wavelet functions is the MRA capability, allowing the simultaneous representation of the neural field at different spatial scales. B-spline scaling and wavelet functions are compactly supported, eliminating  numerical errors caused by the Gaussian basis function. The partition of unity (ability to represent a zero spatial frequency field) of these basis functions enable the reconstruction of the neural field when it saturates. This may be beneficial to model neurological disorders such as epilepsy. Attaining fine-scale approximation of the neural field using Gaussian basis functions results in an ill-posed problem as the number of states is increased significantly. However, this is not the case when using B-spline scaling and wavelet functions due to their semi-orthogonality property.\\ \citet{Freestone2011} showed experimentally that Gaussian functions are a suitable choice for modeling the observation kernel when intracranial measurements are considered. B-splines converge to Gaussian functions as the order of the B-spline, $m$, tends to infinity. The cubic B-spline is already a very close approximation to the Gaussian function. Therefore, these should also be suitable for modeling the observation kernel while having the advantage of being compactly supported.

@@
